---
title: "Preparing RKI COVID-19 Weekly Data for the microsimulation"
date: last-modified
format:
  html:
    toc: true
    number-sections: true
    code-fold: show 
    code-fold-show: true
    code-summary: "Hide code"    
    code-tools: true
    theme: cosmo
---

# Overview
This document processes weekly COVID-19 surveillance data from the Robert Koch Institute (RKI) for use in microsimulation models. It combines weekly incidence data for males and females from 2020–2021, cleans and standardizes the format, and outputs incidence values stratified by age and sex.

# Data Source
All incidence data were obtained from RKI's SurvStat@RKI 2.0 portal: > **Robert Koch Institute**: *SurvStat@RKI 2.0*, https://survstat.rki.de

# Code to process the data

## Initial set-up

Loading libraries, setting template data structure and paths to files. Below is the list of weekly files that will be processed.
```{r initial set-up, message = FALSE, warning = FALSE}
# loading libraries
library(ISOweek)
library(here)
library(readr)
library(DT)
library(lubridate)
library(stringr)
library(purrr)
library(dplyr)

# setting path to files
weekly_path <- here("data", "weekly")

# Constructing a template for the data structure, since not all weekly files
# have the same age groups
template_cols <- c("Sex", 
                   paste0("A", sprintf("%02d", 0:79)), #gives A00:A79
                   "A80+" )

# listing all weekly files in the folder
weekly_files <- list.files(
  weekly_path,
  pattern    = "^SurvStat_\\d{4}_\\d{2}\\.csv$",
  full.names = TRUE
)

datatable(
  data.frame(File = basename(weekly_files)),
  options = list(pageLength = 10, lengthMenu = c(10, 25, 50, 105),
                 searching = FALSE,
                 info      = FALSE),
  rownames = FALSE
)
```

## Making a function to load the data

This function loads the data, cleans the variable names, and extracts the year and week from the filename. 

```{r function to read data}
# creating a function to load and process the files
read_data <- function(path) {
  # read the file
  # skipping the first row, which just has a header
  df <- suppressMessages(
    read_delim(
      path,
      delim   = "\t",           
      locale  = locale(encoding = "UTF-16LE"),  
      trim_ws = TRUE,
      name_repair    = "unique_quiet",
      skip = 1
    ) 
  )
  
  # cleaning variable names
  df <- df %>%
    rename(Sex = 1) %>%
    rename_with(~ substr(.x, 1, 3), -Sex)

  # extract Year and Week from filename
  parts <- str_match(basename(path), "SurvStat_(\\d{4})_(\\d{2})\\.csv")[,2:3]
  
  # returning df with Year and Week added
  df %>%
    mutate(
      Year = as.integer(parts[1]),
      Week = as.integer(parts[2])
    )
}
```

## Loading and processing data

After processing the data I get a data set with the structure of the one shown below

```{r loading and processing data}
# loading in the data
all_weeks <- weekly_files %>%
  set_names() %>%            
  map_dfr(
    read_data,
    .id = "file") 

# renaming one of the columns, and then rearranging the order
all_weeks <- all_weeks %>%
  rename("A80+" = "A80") %>%
  select(Year, Week, all_of(template_cols)) 

# in a few weeks early in 2020 there is no data for females, so 
# I am adding these rows manually
missing_fem <- expand.grid(
  Year = 2020,
  Week = c(1:4, 7),
  Sex  = "female"
)

# converting one week (period of data) into a year
week_in_years <- 1 / 52

# adding the missing female rows and converting all NA to 0
# the incidence rates are number of new cases per 100,000 people during 
# a week. I'm converting this to a yearly rate as micsim works in years 
complete_data <- all_weeks %>%
  bind_rows(missing_fem) %>%
  arrange(Year, Week) %>%
  mutate(across(4:last_col(), ~ if_else(is.na(.), 0, .))) %>%
  mutate(across(4:last_col(), 
                ~ -log(1- (. / 100000))/
                  week_in_years)) 

# adding variables to represent the start and end of the week
complete_data <- complete_data %>%
  mutate(iso_wk1   = sprintf("%04d-W%02d-1", Year, Week),
         iso_wk7   = sprintf("%04d-W%02d-7", Year, Week),
         week_start = ISOweek2date(iso_wk1),
         week_start_num  = decimal_date(week_start),
         week_end = ISOweek2date(iso_wk7),
         week_end_num  = decimal_date(week_end)
  ) %>%
  select(!c(iso_wk1, iso_wk7)) %>%
  relocate(week_start:last_col(), .after = Week) 

# extracting data for males and females
complete_m <- complete_data %>%
  filter(Sex == "male") %>%
  select(week_end_num, "A00":last_col())

complete_f <- complete_data %>%
  filter(Sex == "female") %>%
  select(week_end_num, "A00":last_col())

# print example of what data looks like
datatable(
  complete_data %>% 
    filter(Year == 2021),
  options = list(
    pageLength = 10, lengthMenu = c(10, 25, 50, 105),
    searching = FALSE,
    scrollX  = TRUE,
    info      = FALSE
  ),
  rownames = FALSE
)
```


```{r exporting data}
# exporting data
write_csv(complete_data, file = here("data", "SurvStat_complete_data.csv"))
write_csv(complete_m, file = here("data", "SurvStat_male_data.csv"))
write_csv(complete_f, file = here("data", "SurvStat_female_data.csv"))

saveRDS(complete_data, file = here("data", "SurvStat_complete_data.rds"))
saveRDS(complete_m, file = here("data", "SurvStat_male_data.rds"))
saveRDS(complete_f, file = here("data", "SurvStat_female_data.rds"))
```

## Reproducibility

```{r session-info}
sessionInfo()
```
